---
output:
  pdf_document: default
  html_document: default
---
# Advent Of Code 2022

```{r, echo=FALSE, results='hide'}
knitr::opts_chunk$set(swarning = FALSE, message = FALSE)
rm(list = ls())
```

Some libs and helper functions

```{r}
library(tidyverse)

str2vec <- function(string) {
  return(strsplit(string, "")[[1]])
}
```

## Day 1

### Part 1

Find the Elf carrying the most Calories. How many total Calories is that Elf carrying?

```{r}
raw_data <- read_csv("day1.txt", col_names = FALSE, skip_empty_rows = FALSE)
data <- c()
cumsum <- 0
for (v in raw_data$X1) {
  if (is.na(v)) {
    data <- append(data, cumsum)
    cumsum <- 0
  } else {
    cumsum <- cumsum + v
  }
}

max_cal <- 0
for (v in data) {
  if (v > max_cal) {
    max_cal <- v
  }
}

max_cal
```

### Part 2

Find the top three Elves carrying the most Calories. How many Calories are those Elves carrying in total?

```{r}
max_cals <- c(0, 0, 0)
for (v in data) {
  for (i in 1:3) {
    if (v > max_cals[i]) {
      min_index <- 1
      if (max_cals[2] < max_cals[min_index]) {
        min_index <- 2
      }
      if (max_cals[3] < max_cals[min_index]) {
        min_index <- 3
      }
      max_cals[min_index] <- v
      break
    }
  }
}

sum(max_cals)
```

## Day 2

### Part 1

What would your total score be if everything goes exactly according to your strategy guide?

```{r}
data <- read_delim("day2.txt", col_names = c("You", "Me"), delim = " ") %>%
  mutate(
    You = case_when(
      You == "A" ~ "R",
      You == "B" ~ "P",
      You == "C" ~ "S"
    ),
    Me = case_when(
      Me == "X" ~ "R",
      Me == "Y" ~ "P",
      Me == "Z" ~ "S"
    ),
    ShapeScore = case_when(
      Me == "R" ~ 1,
      Me == "P" ~ 2,
      Me == "S" ~ 3
    ),
    OutcomeScore = case_when(
      Me == You ~ 3,
      Me == "R" & You == "S" ~ 6,
      Me == "P" & You == "R" ~ 6,
      Me == "S" & You == "P" ~ 6,
      TRUE ~ 0
    ),
    Score = ShapeScore + OutcomeScore
  )

sum(data$Score)
```

### Part 2

Following the Elf's instructions for the second column, what would your total score be if everything goes exactly according to your strategy guide?

```{r}
data <- read_delim("day2.txt", col_names = c("You", "Outcome"), delim = " ") %>%
  mutate(
    You = case_when(
      You == "A" ~ "R",
      You == "B" ~ "P",
      You == "C" ~ "S"
    ),
    Outcome = case_when(
      Outcome == "X" ~ "Lose",
      Outcome == "Y" ~ "Draw",
      Outcome == "Z" ~ "Win"
    ),
    OutcomeScore = case_when(
      Outcome == "Lose" ~ 0,
      Outcome == "Draw" ~ 3,
      Outcome == "Win" ~ 6
    ),
    Me = case_when(
      Outcome == "Draw" ~ You,
      Outcome == "Win" & You == "R" ~ "P",
      Outcome == "Win" & You == "P" ~ "S",
      Outcome == "Win" & You == "S" ~ "R",
      Outcome == "Lose" & You == "R" ~ "S",
      Outcome == "Lose" & You == "P" ~ "R",
      Outcome == "Lose" & You == "S" ~ "P",
    ),
    ShapeScore = case_when(
      Me == "R" ~ 1,
      Me == "P" ~ 2,
      Me == "S" ~ 3
    ),
    Score = ShapeScore + OutcomeScore
  )

sum(data$Score)
```

## Day 3

### Part 1

Find the item type that appears in both compartments of each rucksack. What is the sum of the priorities of those item types?

First define a helper funciton to map characters to priority

```{r}
charToPriority <- function(char) {
  ascii <- utf8ToInt(char)
  if (ascii > 90) {
    return(ascii - 96)
  } else {
    return(ascii - 38)
  }
}
```

Then solve

```{r}
common_item.rucksack <- function(items) {
  left <- substr(items, 1, nchar(items) / 2) # raw string
  left <- str2vec(left) # vector of chars
  
  right <- substr(items, (nchar(items) / 2) + 1, nchar(items)) # raw string
  right <- str2vec(right) # vector of chars
  
  for (c in left) {
    if (c %in% right) {
      return(c)
    } 
  }

  return('')
}

data <- read_csv("day3.txt", col_names = c("Items")) %>%
  mutate(
    CommonItem = sapply(Items, common_item.rucksack),
    Priority = sapply(CommonItem, charToPriority)
  )

sum(data$Priority)
```

### Part 2

Find the item type that corresponds to the badges of each three-Elf group. What is the sum of the priorities of those item types?

```{r}
common_item.team <- function(items) {
  items <- strsplit(items, " ")[[1]] # list of list
  rucksack.1 <- str2vec(items[1])
  rucksack.2 <- str2vec(items[2])
  rucksack.3 <- str2vec(items[3])
  for (i in 1:3) {
    rucksack <- str2vec(items[i])
    for (c in rucksack) {
      if (c %in% rucksack.1 & c %in% rucksack.2 & c %in% rucksack.3) {
        return(c)
      }
    }
  }
  return('')
}

data <- read_csv("day3.txt", col_names = c("Items")) %>%
  mutate(TeamId = rep(1:(length(Items) / 3), each = 3)) %>%
  group_by(TeamId) %>%
  summarise(Items = str_c(Items, collapse = " ")) %>%
  mutate(
    CommonItem = sapply(Items, common_item.team),
    Priority = sapply(CommonItem, charToPriority)
  )

sum(data$Priority)
```

## Day 4

### Part 1

In how many assignment pairs does one range fully contain the other?

```{r}
lower_range <- function(r) {
  str_split(r, "-")[[1]][1] %>% as.numeric()
}

upper_range <- function(r) {
  str_split(r, "-")[[1]][2] %>% as.numeric()
}
```

```{r}
data <- read_csv("day4.txt", col_names = c("Elf1", "Elf2")) %>%
  mutate(
    Elf1Start = sapply(Elf1, lower_range),
    Elf1End = sapply(Elf1, upper_range),
    Elf2Start = sapply(Elf2, lower_range),
    Elf2End = sapply(Elf2, upper_range),
    OneContainsAnother = 
      (Elf1Start >= Elf2Start & Elf1End <= Elf2End) | 
      (Elf2Start >= Elf1Start & Elf2End <= Elf1End)
  )

sum(data$OneContainsAnother)
```

### Part 2

In how many assignment pairs do the ranges overlap?

```{r}
data <- read_csv("day4.txt", col_names = c("Elf1", "Elf2")) %>%
  mutate(
    Elf1Start = sapply(Elf1, lower_range),
    Elf1End = sapply(Elf1, upper_range),
    Elf2Start = sapply(Elf2, lower_range),
    Elf2End = sapply(Elf2, upper_range),
    Overlap =
      (Elf2Start >= Elf1Start & Elf1End >= Elf2Start) |
      (Elf1Start >= Elf2Start & Elf2End >= Elf1Start)
  )

sum(data$Overlap)
```

## Day 5

### Part 1

After the rearrangement procedure completes, what crate ends up on top of each stack?

```{r}
# read data as list of string
data <- read_lines("day5.txt")

# find where the data format break is
divider <- which(data == "")

# divide raw data into two parts
raw_stacks <- data[1:(divider - 1)]
raw_instr <- data[(divider + 1):length(data)]

# define function to extract numbers from raw instructions
parse_instr <- function(instr) {
  return(
    instr %>% 
      str_extract("move ([0-9][0-9]*) from ([0-9][0-9]*) to ([0-9][0-9]*)", group = c(1, 2, 3)) %>%
      as.numeric()
  )
}
#####################################
# create dataframe for instructions #
#####################################

instr.orig <- raw_instr %>%
  sapply(parse_instr) %>%
  t() %>%
  as.data.frame() %>%
  rename(NCrates = V1, From = V2, To = V3)

#################################
# create list of vec for stacks #
#################################

# first, convert the stacks to a list of character vectors
stacks.list <- list() 
for (s in raw_stacks) {
  stacks.list <- append(stacks.list, s %>% str2vec() %>% list())
}

# convert the list of character vectors to a matrix of characters
stacks.matrix <- do.call(rbind, stacks.list)

# identify the columns that have actual data in them by 
# isolating the bottom row indexes which are not empty
cols_with_data <- which(stacks.matrix[nrow(stacks.matrix),] != " ")
stacks.matrix <- stacks.matrix[,cols_with_data]

# since the bottom row is that same as indexes, it is
# redundant and can be removed
stacks.matrix <- stacks.matrix[1:nrow(stacks.matrix)-1,]

# mirror the matrix horizontally to work with the rows as left-to-right lists
stacks.matrix <- stacks.matrix[nrow(stacks.matrix):1,]

# convert the matrix back to a list of character vectors
stacks.orig <- stacks.matrix %>% as.data.frame() %>% as.list()

# remove trailing empty characters from the ends of lists
for (i in 1:length(stacks.orig)) {
  stacks.orig[[i]] <- stacks.orig[[i]][stacks.orig[[i]] != " "]
}
```

```{r}
instr <- instr.orig
stacks <- stacks.orig

# apply instrucions to stacks
for (i in 1:nrow(instr)) {
  n_crates <- instr[i,]$NCrates
  from <- instr[i,]$From
  to <- instr[i,]$To
  from_height <- length(stacks[[from]])
  moving_crates <- stacks[[from]][(from_height - n_crates + 1):from_height] %>% rev()
  stacks[[from]] <- stacks[[from]][1:(from_height - n_crates)]
  stacks[[to]] <- append(stacks[[to]], moving_crates)
}

# get top crates
top_crates <- c()
for (i in 1:length(stacks)) {
  top_crates <- append(top_crates, tail(stacks[[i]], 1))
}

# show as one string
top_crates %>% paste(collapse = "")
```

### Part 2

After the rearrangement procedure completes, what crate ends up on top of each stack?

Can be solved the same way but without reversing the crates being moved (see comment)

```{r}
instr <- instr.orig
stacks <- stacks.orig

# apply instrucions to stacks
for (i in 1:nrow(instr)) {
  n_crates <- instr[i,]$NCrates
  from <- instr[i,]$From
  to <- instr[i,]$To
  from_height <- length(stacks[[from]])
  moving_crates <- stacks[[from]][(from_height - n_crates + 1):from_height] # NO REVERSE
  stacks[[from]] <- stacks[[from]][1:(from_height - n_crates)]
  stacks[[to]] <- append(stacks[[to]], moving_crates)
}

# get top crates
top_crates <- c()
for (i in 1:length(stacks)) {
  top_crates <- append(top_crates, tail(stacks[[i]], 1))
}

# show as one string
top_crates %>% paste(collapse = "")
```

## Day 6

### Part 1

How many characters need to be processed before the first start-of-packet marker is detected?

```{r}
data <- read_lines("day6.txt") %>% str2vec()
start_of_packet <- 4
offset <- start_of_packet - 1
marker <- NA
for (i in start_of_packet:length(data)) {
  if (data[(i - offset):i] %>% length() == data[(i - offset):i] %>% unique() %>% length()) {
    marker <- i
    break
  }
}
marker
```

## Part 2

How many characters need to be processed before the first start-of-message marker is detected?

```{r}
data <- read_lines("day6.txt") %>% str2vec()
start_of_message <- 14
offset <- start_of_message - 1
marker <- NA
for (i in start_of_message:length(data)) {
  if (data[(i - offset):i] %>% length() == data[(i - offset):i] %>% unique() %>% length()) {
    marker <- i
    break
  }
}
marker
```












